Geometric Filters for Metric Spaces

### 2018-11-21 paper/draft.tex
Interim

\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{comment}

@@

\newcommand{\parent}{\mathrm{par}}
\newcommand{\point}{p}
\newcommand{\level}{\ell}
\newcommand{\diam}{\mathrm{diam}}
\newcommand{\pointset}{P}
\newcommand{\metricspace}{\mathcal{M}}
\newcommand{\metric}{\delta}
\newcommand{\approxmetric}{\gamma}
\newcommand{\complexity}{C_{\metric}}
\newcommand{\doublingdimension}{\Delta}

@@

\paragraph{Motivation.}
%
Given a set $\pointset:=\{p_1,\ldots,p_n\}$ of $n$ points
in a metric space $(\metricspace,\metric)$, 
consider the following standard operations:

\begin{description}
\item[Nearest Neighbor] Given a point $q\in\metricspace$,
find $p_i\in\pointset$ such that, for all $j=1,\ldots,n$,
\[d(q,p_i)\leq d(q,p_j)\]

\item[Metric approximation] Given $\eps>0$, compute a metric $\approxmetric$
such that for all $1\leq i,j\leq n$,
%
\[\metric(p_i,p_j)\leq\approxmetric(p_i,p_j)\leq (1+\eps)\metric(p_i,p_j).\]

By ``computing'' $\approxmetric$, we mean to compute a $n\times n$-matrix $E$ 
such that $E_{ij}=\approxmetric(p_i,p_j)$.
\end{description}

Assume that an algorithm is given to compute $\metric(\cdot,\cdot)$
for any two points in $\metricspace$ in worst-case complexity $\complexity$.
Then trivially, the above problems can be solved (exactly) in
$O(n\complexity)$ and $O(n^2\complexity)$ time, respectively.
For low-dimensional\footnote{we will make this notion precise
later in the text} metric spaces, we can improve these bounds:
Using cover trees~\cite{cover-trees}, we can preprocess the point set
in $O(\complexity n\log n)$ time to answer nearest neighbor queries
in $O(\complexity \log n)$ time. Combining cover trees with well-separated
pair decompositions (WSPD), we obtain an $\eps$-spanner for the graph
induced by $\metric$ in $O(\complexity n\log n)$ time and
of linear size (assuming, for simplicity, that $\eps$ is a constant).
Out of this spanner,
we can construct an approximate metric
with an all-pair shortest path computation,
resulting in a total running time of
\[O(\complexity n\log n+ n^2\log n)\]
using Johnson's algorithm~\cite{johnson}.

In the aforementioned techniques, $\complexity$ is usually considered
as a constant-time operation and hence ignored in the analysis.
There are good reasons for that: the most common case of a metric space
is $\metricspace=\R^d$ with $d$ some constant, in which case $\metric$ can be evaluated in $O(d)=O(1)$ time.
Even if $d$ is considered non-constant,
it can always be assumed that $d\leq n$, hence $\complexity$ is at most $O(n)$.
Another typical assumption is that all pairwise distances are part of the input
in which case $\complexity$ is $O(1)$.

However, we argue that in some situations, distance computations
in $\metricspace$ can be costly and $\complexity$ might be incomparable
with $n$. Our motivating examples come from topological summaries
such as persistence diagrams or Reeb graphs, which are of interest
in the field of topological data analysis. A persistence diagram
is a point set in $\R^2$, and the distance between two diagrams
is determined by a min-cost matching between the point sets.
If the diagrams have $N$ points, computing this matching requires
polynomial time in $N$, and $N$ might well be larger than $n$, the number
of diagrams considered. For the case of Reeb graphs, the situation is even
worse: while several metrics on Reeb graphs have been proposed,
not even an constant-factor approximation algorithm is known that runs
in polynomial time in the size of the graphs.

@@

\paragraph{Contribution.}
%
We demonstrate that the number of distance computations can be further
reduced by a careful re-implementation of the underlying geometric algorithms.
Our improvement is based on the following simple observation: many geometric
constructions (such as cover tree and WSPD constructions) require distance
computations only in the form of the following predicate: 
``Is $\metric(p_i,p_j)\leq t$'', where $t$ is a real value.
We show that using the information of the already computed distances and
the triangle inequality, we can answer the above predicate in many cases
without computing the actual distance.

We describe our idea with an example. In Figure~\ref{fig:1st_example},
we see a weighted graph on $P$ whose edges are the distances that have
been computed so far; note that $\metric(p_1,p_2)$ is unknown. 
The shortest path from $p_1$ to $p_2$ has length $9$, which clearly
constitutes an upper bound on the distance by triangle inequality.
However, we can also infer that $\metric(p_1,p_2)\geq 3$:
otherwise, the path from $p_3$ to $p_4$ via $p_1$ and $p_2$
would be shorter than the edge $(p_3,p_4)$, again contradicting
triangle inequality.
Hence, the question whether $\metric(p_1,p_2)\leq t$ can be answered positively
for $t>=9$ and negatively for $t<3$. Only if $t\in[3,9)$, the actual
distance computation is necessary. This approach is reminiscent to the usage
of \emph{numerical filters} in geometric computing in the CGAL library,
where geometric predicates in exact number types are first evaluated using 
certified floating-point arithmetic, and exact computation is only
invoked if the numerical evaluation did not arrive at a decisive result
(e.g.~\cite{bbp-interval,kerber-phd}).

@@

In our implementation, we maintain the best lower and upper bounds of all
pairwise distances in an $n\times n$ matrix of intervals. We show how to
update this matrix in $O(n^2)$ time when a new edge is added.
%We point out once more that a quadratic update time only pays off
%if distance computations are expensive, say $\complexity=\Omega(n^2)$.

@@

We also provide experimental results for the case of persistence diagrams
which motivated this line of research. We point out that the space of
persistence diagrams fails to be ``low-dimensional''. To be precise,
its doubling dimension is $\infty$, and a finite set of $n$ diagram
can have the worst-case doublng dimension $\log n$. This worst-case
behavior is also the expected one if the diagrams are chosen randomly
(in a sense made precise in Section~\ref{sec:metric_spaces_top_summaries}). 
However, we demonstrate that
for natural collections of persistence diagrams, a substantial fraction
of distance computations can be avoided. In detail, $\ldots$

@@

@@

We revise the concepts that were designed for nearest neighbor queries
and metric approximation in metric spaces with low dimensionality.
We will also review the computation methods and how they depend
on the predicate ``$\metric(p_i,p_j)\leq t$?'' mentioned above.

@@

\section{Distance Bounds in Partially Revealed Metrics}
%
We describe our algorithm to maintain lower and upper bounds for the
distance of any two points in our (finite) metric space $(P=\{p_1,\ldots,p_n\},\metric)$. We assume for simplicity that all points in $P$ are distinct.

We maintain an $n\times n$ matrix $M$ whose entries
are intervals $M_{ij}=[a_{ij},b_{ij}]$, meaning that the distance
of $p_i$ and $p_j$ lies in that interval. 
We initialize $M_{ij}=[0,\infty]$ for all $i\neq j$
and $M_{ij}=[0,0]$ for $i=j$.
We describe next how to update $M$ when a new distance has been computed.

Note that an explicit graph representation of the partially revealed metric,
such as in Figure~\ref{fig:1st_example} is not necessary. However, for 
the sake of clarity,
this graph is helpful. So, let $G=(P,E)$ be the undirected weighted graph
where an edge between $p_i$ and $p_j$ with weight $\metric(p_i,p_j)$
exists if that distance has been computed. 
We maintain the invariant that the upper bound $b_{ij}$ equals the 
shortest path from $p_i$ to $p_j$ in $G$.

Assume that $\metric(p_i,p_j)=v\in\R$ has been computed during the algorithm.
First, we reset $M_{i,j}$ and $M_{j,i}$ to $[v,v]$. 
In $G$, the edge $(p_i,p_j)$ is added with weight $v$.
To update the upper bound of some entry $M_{k,\ell}$,
we observe that the shortest path from $p_k$ to $p_\ell$ might now
go through the new edge. Hence, we update
%
\[b_{k,\ell}\gets \min\{b_{k,\ell},b_{k,i}+v+b_{j,\ell},b_{k,j}+v+b_{i,\ell}\}\]
%
Repeating this for all $k,\ell$ yields the updated upper bounds in $O(n^2)$
time.

For the lower bound, we observe that for any $1\leq k,\ell\leq n$,
%
\[v-b_{k,i}-b_{\ell,j}\]
%
is a lower bound for $\metric(p_k,p_\ell)$. Indeed, this follows from
the triangle inequality
%
\[\metric(p_i,p_j)\leq \metric(p_i,p_k)+\metric(p_k,p_\ell)+\metric(p_\ell,p_j)\]
by rearranging terms and plugging in the upper bounds for $\metric(p_i,p_k)$
and $\metric(p_k,p_\ell)$. An analogue bound holds for the roles of $i$
and $j$ swapped. Hence, we update
%
\[a_{k,\ell}\gets \max\{a_{k,\ell},v-b_{k,i}-b_{\ell,j},v-b_{k,j}-b_{\ell,i}\}\]
%
for all $k,\ell$ and obtain the updated $M$ in $O(n^2)$ time.

\medskip

With the matrix $M$, we can define filtered versions of our geometric
algorithms very easily: whenever the algorithm makes a query of the form
``Is $\metric(p_i,p_j)\leq t$?'', it is checked whether $a_{ij}> t$,
in which case the answer is ``yes'', or whether $b_{ij}\leq t$,
in which case the answer is ``no''. If none of the cases applies, 
$v\gets\metric(p_i,p_j)$ is computed, 
the matrix $M$ is updated as described above, and the predicate is answered
by comparing $v$ and $t$.

\section{Experiments on Point Clouds}
\label{sec:experiments_point_cloud}

\section{Metric spaces of Topological Summaries}
\label{sec:metric_spaces_top_summaries}

\section{Experiments on Persistence Diagrams}
\label{sec:experiments_persistence_diagrams}

>>>>

\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{comment}
\usepackage{braket}

@@

\newcommand{\parent}{\mathrm{par}}
\newcommand{\point}{p}
\newcommand{\level}{\ell}
\newcommand{\diam}{\mathrm{diam}}
\newcommand{\pointset}{P}
\newcommand{\distspace}{\mathcal{M}}
\newcommand{\dist}{\delta}
\newcommand{\adist}{\gamma}
\newcommand{\complexity}{C_{\dist}}
\newcommand{\doublingdimension}{\Delta}

@@

\paragraph{Motivation.}
%
Given a set $\pointset:=\{p_1,\ldots,p_n\}$ of $n$ points
in a metric space $(\distspace,\dist)$, 
consider the following standard operations:

\begin{description}
\item[Nearest Neighbor] Given a point $q\in\distspace$,
find $p_i\in\pointset$ such that, for all $j=1,\ldots,n$,
\[d(q,p_i)\leq d(q,p_j)\]

\item[Metric approximation] Given $\eps>0$, compute a metric $\adist$
such that for all $1\leq i,j\leq n$,
%
\[\dist(p_i,p_j)\leq \adist(p_i,p_j) \leq (1+\eps) \dist(p_i,p_j).\]

By ``computing'' $\adist$, we mean to compute a $n\times n$-matrix $E$ 
such that $E_{ij}=\adist(p_i,p_j)$.
\end{description}

Assume that an algorithm is given to compute $\dist(\cdot,\cdot)$
for any two points in $\distspace$ in worst-case complexity $\complexity$.
Then trivially, the above problems can be solved (exactly) in
$O(n\complexity)$ and $O(n^2\complexity)$ time, respectively.
For low-dimensional\footnote{we will make this notion precise
later in the text} metric spaces, we can improve these bounds:
Using cover trees~\cite{cover-trees}, we can preprocess the point set
in $O(\complexity n\log n)$ time to answer nearest neighbor queries
in $O(\complexity \log n)$ time. Combining cover trees with well-separated
pair decompositions (WSPD), we obtain an $\eps$-spanner for the graph
induced by $\dist$ in $O(\complexity n\log n)$ time and
of linear size (assuming, for simplicity, that $\eps$ is a constant).
Out of this spanner,
we can construct an approximate metric
with an all-pair shortest path computation,
resulting in a total running time of
\[O(\complexity n\log n+ n^2\log n)\]
using Johnson's algorithm~\cite{johnson}.

In the aforementioned techniques, $\complexity$ is usually considered
as a constant-time operation and hence ignored in the analysis.
There are good reasons for that: the most common case of a metric space
is $\distspace=\R^d$ with $d$ some constant, in which case $\dist$ can be evaluated in $O(d)=O(1)$ time.
Even if $d$ is considered non-constant,
it can always be assumed that $d\leq n$, hence $\complexity$ is at most $O(n)$.
Another typical assumption is that all pairwise distances are part of the input
in which case $\complexity$ is $O(1)$.

However, we argue that in some situations, distance computations
in $\distspace$ can be costly and $\complexity$ might be incomparable
with $n$. Our motivating examples come from topological summaries
such as persistence diagrams or Reeb graphs, which are of interest
in the field of topological data analysis. A persistence diagram
is a point set in $\R^2$, and the distance between two diagrams
is determined by a min-cost matching between the point sets.
If the diagrams have $N$ points, computing this matching requires
polynomial time in $N$, and $N$ might well be larger than $n$, the number
of diagrams considered. For the case of Reeb graphs, the situation is even
worse: while several metrics on Reeb graphs have been proposed,
not even an constant-factor approximation algorithm is known that runs
in polynomial time in the size of the graphs.

@@

\paragraph{Contribution.}
%
We demonstrate that the number of distance computations can be further
reduced by a careful re-implementation of the underlying geometric algorithms.
Our improvement is based on the following simple observation: many geometric
constructions (such as cover tree and WSPD constructions) require distance
computations only in the form of the following predicate: 
``Is $\dist(p_i,p_j)\leq t$'', where $t$ is a real value.
We show that using the information of the already computed distances and
the triangle inequality, we can answer the above predicate in many cases
without computing the actual distance.

We describe our idea with an example. In Figure~\ref{fig:1st_example},
we see a weighted graph on $P$ whose edges are the distances that have
been computed so far; note that $\dist(p_1,p_2)$ is unknown. 
The shortest path from $p_1$ to $p_2$ has length $9$, which clearly
constitutes an upper bound on the distance by triangle inequality.
However, we can also infer that $\dist(p_1,p_2)\geq 3$:
otherwise, the path from $p_3$ to $p_4$ via $p_1$ and $p_2$
would be shorter than the edge $(p_3,p_4)$, again contradicting
triangle inequality.
Hence, the question whether $\dist(p_1,p_2)\leq t$ can be answered positively
for $t>=9$ and negatively for $t<3$. Only if $t\in[3,9)$, the actual
distance computation is necessary. This approach is reminiscent to the usage
of \emph{numerical filters} in geometric computing in the CGAL library,
where geometric predicates in exact number types are first evaluated using 
certified floating-point arithmetic, and exact computation is only
invoked if the numerical evaluation did not arrive at a decisive result
(e.g.~\cite{bbp-interval,kerber-phd}).

@@

In our implementation, we maintain the best lower and upper bounds of all
pairwise distances in an $n\times n$ matrix of intervals. We show how to
update this matrix in $O(n^4)$ time when a new edge is added.
%We point out once more that a quadratic update time only pays off
%if distance computations are expensive, say $\complexity=\Omega(n^2)$.

@@

We also provide some experimental results for the case of persistence diagrams
which motivated this line of research. We point out that the space of
persistence diagrams fails to be ``low-dimensional''. To be precise,
its doubling dimension is $\infty$, and a finite set of $n$ diagram
can have the worst-case doublng dimension $\log n$. This worst-case
behavior is also the expected one if the diagrams are chosen randomly
(in a sense made precise in Section~\ref{sec:metric_spaces_top_summaries}). 
However, we demonstrate that
for natural collections of persistence diagrams, a substantial fraction
of distance computations can be avoided. In detail, $\ldots$

@@

\section{Metric Approximation}

It is known that asymptotically optimal spanner can be constructed by the following greedy algorithm: 
start with an empty graph on vertices $p_i$,
sort all edges $\{p_i, p_j\}$ by the length $\dist(p_i, p_j)$ in ascending order and keep adding them to the graph
until it becomes a $1+\eps$-spanner. In our setting this is the most expensive approach, because it requires
computing all pairwise distances first.

 In our setting this is the most expensive approach, because it requires
 computing all pairwise distances first. We propose an alternative approach, which we call
 \textit{blind} spanner, meaning that the algorithm does not have access to $\dist(p_i, p_j)$
 before it decides to add the edge $(p_i, p_j)$. 
 The algorithm also maintains lower and upper bounds
 for all pairwise distances, which are updated after a new edge is added. and the stopping criterion
 is based on these bounds, namely, if $a_{i, j}$ and $b_{i,j}$ denote lower and upper bounds
 for the distance between $p_i$ and $p_j$, respectively, then the algorithm stops, if $b_{i,j} / a_{i, j} \leq 1 + \eps$
 for all $i \neq j$.

\begin{algorithmic}
\label{alg:blind_spanner}
\Function{BlindSpanner}{$P, \eps$}
    \State {$E \gets \emptyset$}
    \State {$a_{i,j} \gets 0$ for all $1 \leq i,j \leq n$}
    \State {$b_{i,j} \gets \infty$ for all $1 \leq i,j \leq n, i \neq j$}
    \While {$\exists i \neq j : b_{i,j} / a_{i,j} > 1 + \eps$}
    \State {$(i,j) \gets GetEdgeToAdd$}
    \State {$v \gets \dist(p_i, pj)$}
    \State {Add weighted edge $(p_i, p_j, v)$ to $E$}
    \State {$UpdateBounds(i, j, v)$}
    \EndWhile
\EndFunction
\end{algorithmic}

In this pseudocode we adopt the convention $ \textrm{positive number} / 0 = \infty$ and $\infty > \textrm{any real number}$,
thus making the predicate in the while loop well-defined.  The procedure $UpdateBounds$ is desribed in the next subsection, 
now we only mentioning that it updates all upper and lower bounds based on triangle inequality involving the edge $(p_i, p_j)$,
and the update is monotonic: no upper bound can increase, no lower bound can decrease (in particular, $a_{i,j} = b_{i,j} = v$).
The most interesting detail of this algorithm is the way of choosing the next edge to be added, and here 
there are several possibilities, which give very different results.

The first possibility will be called $BlindGreedy$, where we choose a pair $p_i, p_j$ that maximizes
the ratio $b_{i,j} / a_{i,j}$. If the number of maximizers is more than 1, we choose one of them at random.
Note that our conventions imply that in $BlindGreedy$ the edges that have $a_{i,j} = 0$ or $b_{i,j} = \infty$
have the highest priority, so the algorithm first ensures that the graph is connected and there are positive
lower bounds for every edge before it will start adding any other edges.

The second possibility will be referred to as $BlindRandom$, where among all edges $(p_i, p_j)$
for which $b_{i,j} / a_{i,j} > 1 + \eps$ we choose one uniformly at random. We may vary this approach

 \paragraph{Maintaining lower and upper bounds}
 More specifically,
 we maintain an $n\times n$ matrix $M$ whose entries
are intervals $M_{ij}=[a_{ij},b_{ij}]$, meaning that the distance
between $p_i$ and $p_j$ lies in that interval. 
We initialize $M_{ij}=[0,\infty]$ for all $i\neq j$
and $M_{ij}=[0,0]$ for $i=j$.  Assume that $\dist(p_i,p_j)=v\in\R$ has been computed.
First, we reset $M_{i,j}$ and $M_{j,i}$ to $[v,v]$. 
To update the upper bound of some entry $M_{k,\ell}$,
we observe that the shortest path from $p_k$ to $p_\ell$ might now
go through the new edge. Hence, we update
%
\[b_{k,\ell}\gets \min_{i,j}\{b_{k,\ell},b_{k,i}+v+b_{j,\ell},b_{k,j}+v+b_{i,\ell}\}\]
%
Repeating this for all $k,\ell$ yields the updated upper bounds in $O(n^2)$
time per entry.

For the lower bound, we observe that for any $1\leq k,\ell\leq n$,
%
\[v-b_{k,i}-b_{\ell,j}\]
%
is a lower bound for $\dist(p_k,p_\ell)$. Indeed, this follows from
the triangle inequality
%
\[\dist(p_i,p_j)\leq \dist(p_i,p_k)+\dist(p_k,p_\ell)+\dist(p_\ell,p_j)\]
by rearranging terms and plugging in the upper bounds for $\dist(p_i,p_k)$
and $\dist(p_k,p_\ell)$. An analogue bound holds with $i$ and $j$ swapped. Hence, we update
%
\[a_{k,\ell}\gets \max\{a_{k,\ell},v-b_{k,i}-b_{\ell,j},v-b_{k,j}-b_{\ell,i}\}\]
%
for all $k,\ell$ and obtain the updated $M$ in $O(n^4)$ time.

\section{Approximate Nearest Neighbor}

In this section we consider the standard problem of finding an approximate nearest neighbor: given
$n$ points $P = \Set{p_1, \dots, p_n}$, a query point $q$ and a real number $\eps > 0$,
find $p_i$ such that $\dist(p_i, q) \leq (1 + \eps) \min_{k} \dist(q, p_k)$. We assume for simplicity
that all exact pairwise distances $\dist(p_i, p_j)$ are already computed.

\begin{lemma}
    Let $p_i, p_j \in P$ be such that $\dist(p_j, q) \geq \frac{\dist(p_i, q)}{1 + \eps}$,
    and denote $r = \dist(p_i, q) / (1 + \eps)$, $r_j = \dist(p_j, q)$.
    Then for every $x \in B(p_j, r - r_j)$ we have $\dist(x, q) \geq r$
    and for every $y \notin B(p_j, r + r_j)$ we have $\dist(y, q) \geq r$.
\end{lemma}
Proof. The triangle inequality.

\begin{lemma}
\end{lemma}

The algorithm can be informally described as follows. Pick a random candidate $p_i$ and try to verify that
it is an approximate nearest neighbor by scanning all other points. Pick the next random point $p_j$
and assume that $\dist(p_j, q) \geq \frac{\dist(p_i, q)}{1 + \eps}$. This allows us to discard not just the point
$p_j$, but also the points that are very close to $p_j$ and very far from $p_j$ using triangle inequality.
Indeed, consider the ball of radius $ \dist(p_j, q) - \frac{ \dist(p_i, q) }{ 1+\eps} $ centered at $p_j$.
The triangle inequality implies that all points in this ball are at the distance at least $\frac{ \dist(p_i, q) }{ 1+\eps}$
from $q$. Now consider the ball of radius $\dist(p_j, q) + 2 \frac{\dist(p_i, q)}{1+ \eps}$ centered at $p_j$. 
Every point outside of this ball is a

The algorithm that we analyze is based on the following simple observation.
Suppose that we want to to prove that a candidate for the approximate nearest neighbor is $p_1$,

\begin{algorithmic}
\Function{ApproximateNearestNeighbor}{$P, q, \eps$}
\For {$i \in \{\,1, \dots, |P|\,\}$}
\State {$a_i \gets 0$}
\State {$b_i \gets \infty$}
\EndFor
\State {$i_0 \gets \argmin u $}
\State {$B  \gets \{\, i \mid l_i < u[i_0]\,\}$}
\If {$B = \emptyset$}
    \State {\Return {$p_{i_0}$}}
\Else
    \State{$j \gets \mbox{ random element of } C$}
    \State{$u_j \gets d(p_j, q)$}
    \State{$l[j] \gets d(p_j, q)$}
    \State{Mark distance from $q$ to $p_j$ as computed}
    \State{\Call{UpdateBounds}{$l, u, d(p_j, q), j$}}
\EndIf
\EndFunction
\end{algorithmic}

The goal is to compute
as few $\dist(q, p_k)$ as possible.

@@

We revise the concepts that were designed for nearest neighbor queries
and metric approximation in metric spaces with low dimensionality.
We will also review the computation methods and how they depend
on the predicate ``$\dist(p_i,p_j)\leq t$?'' mentioned above.

@@
